Substructure Boosts to Dark Matter Annihilation from Sommerfeld Enhancement

### 2009-02-19 tex/somm.tex
Added VL section

\usepackage{graphicx,epsfig}
\usepackage{amsmath,amssymb}
\usepackage{subfigure,epsfig}
%\usepackage{feynmp}
\usepackage{float}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `basename #1 .tif`.png}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `basename #1 .tif`.png}

@@

\begin{figure*}[t]
\includegraphics[clip=]{2dboost8_65_-0.7.eps}%%BoundingBox: 178 291 400 474
\includegraphics[clip=]{2dboost8_65_-1.0.eps}\\%%BoundingBox: 190 291 430 474
%\includegraphics[clip=]{2dboost8_65_-3.eps}%%BoundingBox: 178 291 400 474
%\includegraphics[clip=]{2dboost8_65_-12.eps}%%BoundingBox: 190 291 430 474
\caption{The total boost for a $M=10^8$ \Msun\ for different slopes
$n$ for the subhalo mass function (\emph{top panels}) and for
different values of the dark matter thermal free streaming limit
$m_0$ (\emph{bottom panels}).}%
\label{fig:subboostslope}%
\end{figure*}

@@

The effect of the thermal free-streaming scale of \DM\ on the
substructure boost is shown in Figure \ref{fig:subboostslope}
(\emph{bottom panels}). This effect is relatively large since the
free-streaming scale sets the scale of the smallest and most
concentrated \DM\ halos and raising or lowering this scale therefore
changes an important part of the contribution to the substructure
boost. For very large values of the free-streaming scale such as $m_0
= 10^{-3}$ \Msun, the substructure boost becomes relatively small
BOVY: BE MORE PRECISE HERE. When the free-streaming scale becomes very
small, much of the substructure is in very small, cold clumps, such
that the total boost becomes very large. BOVY: BE MORE PRECISE HERE.

BOVY: Also describe different thermal free streaming limits here?

@@

\section{The annihilation signal from subhalos in \VL\ 2}

\VL 2 data \cite{2008Natur.454..735D} \footnote{Available at
\texttt{http://www.ucolick.org/\urltilda$\!$diemand/vl/data.html}}.

\begin{figure}
\centering
\includegraphics{hist_-2_-3_10_1.eps}
\caption{The distances to and masses of subhalos within the tidal
radius of the main halo in \VL\ 2. The distances are for three random
observers at a distance of 8.5 kpc from the galactic center of the
main halo.}%\
\label{fig:VL23_1}%
\end{figure}

\begin{figure}
\centering
\includegraphics{hist_-2_-3_10_2.eps}
\caption{Left: Histogram of the predicted gamma-ray flux coming from
indivual subhalos contained within the tidal radius of the main halo
in \VL\ 2 for three random observer placed at a distance of 8.5 kpc
from the center of the main \VL\ halo for \mv/\mdm\ = 10$^{-2}$ and
$\alpha = 10^{-3}$. Right: Histogram of the total boosts of the
subhalos for the same values of \mv/\mdm\ and $\alpha$.}%\
\label{fig:VL23_2}%
\end{figure}

\begin{figure}
\centering
\includegraphics{hist_-2_-3_10_3.eps}
\caption{Total gamma-ray flux from subhalos in \VL\ 2 for 1,000 random
observers at 8.5 kpc from the center of the main \VL\ halo for
\mv/\mdm\ = 10$^{-2}$ and $\alpha = 10^{-3}$.}%\
\label{fig:VL23_3}%
\end{figure}

@@

@@

>>>>

\usepackage{graphicx,epsfig}
\usepackage{amsmath,amssymb}
%\usepackage{float}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `basename #1 .tif`.png}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `basename #1 .tif`.png}

@@

\begin{figure*}[t]
\includegraphics[clip=]{2dboost8_65_-0.7.eps}%%BoundingBox: 178 291 400 474
\includegraphics[clip=]{2dboost8_65_-1.0.eps}\\%%BoundingBox: 190 291 430 474
\includegraphics[clip=]{2dboost8_65_-3.eps}%%BoundingBox: 178 291 400 474
\includegraphics[clip=]{2dboost8_65_-12.eps}%%BoundingBox: 190 291 430 474
\caption{The total boost for a $M=10^8$ \Msun\ for different slopes
$n$ for the subhalo mass function (\emph{top panels}) and for
different values of the dark matter thermal free streaming limit
$m_0$ (\emph{bottom panels}).}%
\label{fig:subboostslope}%
\end{figure*}

@@

The effect of the thermal free-streaming scale of \DM\ on the
substructure boost is shown in Figure \ref{fig:subboostslope}
(\emph{bottom panels}). This effect is relatively large since the
free-streaming scale sets the scale of the smallest and most
concentrated \DM\ halos and raising or lowering this scale therefore
changes an important part of the contribution to the substructure
boost. For very large values of the free-streaming scale such as $m_0
= 10^{-3}$ \Msun, the substructure boost becomes about an order of
magnitude smaller. When the free-streaming scale becomes very
small, much of the substructure is in very small, cold clumps, such
that the total boost becomes about an order of magnitude larger.

@@

\begin{figure}
\centering
\includegraphics[clip=]{hist_-2_-3_10_1.eps}
\caption{The distances to and masses of subhalos within the tidal
radius of the main halo in \VL\ II. The distances are for three random
observers at a distance of 8.5 kpc from the galactic center of the
main halo.}%\
\label{fig:VL23_1}%
\end{figure}

\begin{figure}
\centering
\includegraphics{hist_-2_-3_10_2.eps}
\caption{Left: Histogram of the predicted gamma-ray flux coming from
indivual subhalos contained within the tidal radius of the main halo
in \VL\ II for three random observer placed at a distance of 8.5 kpc
from the center of the main \VL\ halo for \mv/\mdm\ = 10$^{-2}$ and
$\alpha = 10^{-3}$. Right: Histogram of the total boosts of the
subhalos for the same values of \mv/\mdm\ and $\alpha$.}%\
\label{fig:VL23_2}%
\end{figure}

\begin{figure}
\centering
\includegraphics{hist_-2_-3_10_3.eps}
\caption{Total gamma-ray flux from subhalos in \VL\ II for 1,000 random
observers at 8.5 kpc from the center of the main \VL\ halo for
\mv/\mdm\ = 10$^{-2}$ and $\alpha = 10^{-3}$.}%\
\label{fig:VL23_3}%
\end{figure}

@@

\section{The annihilation signal from subhalos in \VL\ II}

Besides the satellite galaxies that have been observed in the Galactic
halo \CDM\ predicts that the halo is filled with hundreds of
individual subhalos ---a fact which gives rise to the so-called
``missing sattelites problem''
\cite{1999ApJ...522...82K,1999ApJ...524L..19M}. Each of these subhalos
has a mass of the order of the masses of the observed \dSphs\ or
smaller, and thus will be kinematically cold with respect to the main
halo. Thus, the total \DM\ annihilation signal coming from all of the
substructure will be boosted by the Sommerfeld enhancement described
above, such that an appreciable \DM\ annihilation signal could be
expected from all of the substructure. In this section we get a
feeling for the expected signal from \DM\ substructures in a typical
Galaxy-sized halo. If the \DM\ annihilation signal from subhalos
without star formation could be detected this could give a clear
signal that the \DM\ exists and would open up a whole new arena for
testing \CDM.

@@

We use the distribution of \DM\ subhalos from the \VL\ II numerical
$N$-body simulation\footnote{Available at
\texttt{http://www.ucolick.org/\urltilda$\!$diemand/vl/data.html}}
\cite{2008Natur.454..735D}. \VL\ II is a state-of-the-art $N$-body
simulation of a Galaxy-sized halo since redshift 104.3. It sampled the
region of the main halo with 1,1 $\times$ 10$^9$ particles, each with
a mass of 4,100 \Msun. Over 40,000 subhalos can be resolved within 402
kpc of the center of the main hala and 20,047 subhalos are detected
which had a peak circular velocity larger than 4 km s$^{-1}$ at some
time. The properties of the subhalos in this latter sample can be
accurately established, \ie, they are unaffected by resolution
effects. We select from this sample the 9,830 subhalos which lie
within the tidal radius of the main halo, which is approximately 462
kpc. We show the distribution of these subhalos in mass and distance
from the Sun in Figure \ref{fig:VL23_1}. The have placed the Sun at
random locations at a distance of 8.5 kpc of the center of the main
halo, corresponding to a earth-based-observer-like vantage point. The
distribution of the distances in Figure \ref{fig:VL23_1} is given for
three of these random observers and the distribution is very similar
for all three observers. We see that most of the subhalos have masses
in the range 10$^4$ -- 10$^6$ \Msun, with subhalos with masses of the
order of the \dSphs\ being relatively rare. The lower limit is of
course set by the finite numerical resolution of the simulation and
doesn't correspond to a physical cut-off ---indeed, the physical
cut-off lies at the thermal free-streaming scale. The distribution of
the distance to the subhalos shows that the distribution of the
subhalos is strongly radially anti-biased, with a sharp cut-off due to
our cut at the tidal radius of the main halo. This radial anti-bias is
due to the fact that there is much more space far away from the center
of the galaxy, and because of the strong tidal forces near the center
of the galaxy, which strongly affects halos which pass near the center
of the galaxy.

For each halo we can estimate the velocity dispersion using the
mass-velocity dispersion relation given in \eqnname\
\ref{eq:vdispscaling} and we can calculate the characteristic density
\rhos\ and the characteristic length-scale \rs\ from the values of
\vmax\ and \rmax\ given for each subhalo, by the procedure outlined at
the end of Sec.~\ref{sec:DMprofile}. For a given set of particle
physics parameters \mv/\mdm\ and $\alpha$ this allows us to calculate
the total boost to the \DM\ annihilation cross section for each of the
subhalos, and by using the assumptions about the particle physics
given in Sec.~\ref{sec:partphys} we can also calculate the photon flux
from \DM\ annihilation coming from each subhalo, and the total
gamma-ray flux coming from all of the subhalos combined.

We will only cover some typical cases for the Sommerfeld enhancement
here which suffice to show the general trend. We consider the particle
physics parameters \mv/\mdm\ and $\alpha$ which give rise to the
smallest Sommerfeld enahcements in Figures \ref{fig:VL23_2} and
\ref{fig:VL23_3}, \ie, \mv/\mdm\ = 10$^{-2}$ and $\alpha$ = 0.001. The
left panel of Figure \ref{fig:VL23_2} shows the distribution of
gamma-ray fluxes from \DM\ annihilation for the subhalos, while the
right panel shows the distribution of the total boost factors. As can
be seen, the total boost factors are quite small: the typical boost
factor is only about 30 with the highest boost factors being about
100. The total gamma-ray flux coming from subhalos is then also quite
small in this case, as is shown in Figure \ref{fig:VL23_3}, in which
the total gamma-ray flux is shown for 1,000 random observer at a
distance of 8.5 kpc from the galactic center. At most observation
locations this total flux is about 10$^{-7}$ cm$^2$ s$^{-1}$, with the
high-end tail not extending further than about half an order of
magnitude from this. This is still a gamma-ray flux which is quite
large, showing that the cumulative signal from all of the subhalos is
dominated by the contribution from the subhalos which give rise to the
largest gamma-ray fluxes.

Figures \ref{fig:VL51_1} and \ref{fig:VL51_2} show the same for a
different extremum of the Sommerfeld enhancement: the case in which
\mv/\mdm\ = 10$^{-5}$ and $\alpha$ = 0.1. As we saw before, in this
case the Sommerfeld enhancement can reach very high values, and thus
we see that the typical \DM\ annihilation signal coming from \DM\
subhalos is large: the typical flux for one subhalo is equal to the
total flux in the scenario described in the previous paragraph. The
total boost factors for most of the halos are very large, of the order
of 10$^7$, which gives rise to this very large photon-flux. The total
predicted flux in this case is many orders of magnitude larger than
the total flux in the other extremum of the Sommerfeld
enhancement. Thus, the exact value of the parameters \mv/\mdm\ and
$\alpha$, upon which the Sommerfeld enhancement depends, matters a
great deal for the predicted gamma-ray flux.

Finally, we look at the expected signal coming from subhalos when the
Sommerfeld enhancement is at a resonance, the same distribution as
before are shown in Figures \ref{fig:VLresonance1} and
\ref{fig:VLresonance2}. Again we find large boost factors and
individual and total gamma-ray fluxes, although they are smaller than
the values found for the \mv/\mdm\ = 10$^{-5}$ and $\alpha$ = 0.1
parameter set. However, since these values are near a resonance in a
region of parameter space around which the predicted signal looks more
like that for the parameter set \mv/\mdm\ = 10$^{-2}$ and $\alpha$ =
0.001, this again shows that whether dark halos, \ie, subhalos without
star formation, can be detected through \DM\ annihilation products
depends very sensitively on the particle physics parameters upon which
the Sommerfeld enhancement depends.

In conclusion we can say that in a scenario in which there is
Sommerfeld enhancement there is much more hope to detect dark subhalos
than in more conventional scenarios, but that the details of the
prediction depend strongly on the underlying particle physics.
